---
title: "Using trekR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using trekR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


```



```{r message = FALSE}
library(trekR)
library(lubridate)
library(dplyr)
```

## Dataset
The `deer` data set contains GPS data (in UTMs) for a three individuals in Utah. 
```{r data }
data(deer)
head(deer)

str(deer)
```

## Setting Up the Data
The example shown below is one of many ways the data could be processed. The important components required is a `list` object containing home range kernels for each interval that is created for at least one individual. 

For example, for a single individual if I wanted to break a one month period into 10-day intervals I would get a total of three 10-day intervals. These intervals should be in a `list` object, and each element within the `list` should represent a single interval. 

### Create Intervals
We created 10-day intervals for deer in the January and July. We first created a `new` column using the `floor_Date()` function to obtain the 10-day extents within the months. We included the 31st day (if a month had 31 days) in the third 10-day interval to prevent the creation of a 4th 10-day interval with only a single day in it.   

### Intervals for January
```{r Jan Intervals}
jan <- deer %>%
  # Creates a new column assigning the first day in the 10-day interval in which
  # the date falls under (e.g., 01-03-2021 would be in the first 10-day interval
  # so the `floor_date` assigned to it would be 01-01-2021)
  mutate(new = floor_date(date, "10 days")) %>%
  # For any months that has 31 days, the 31st day would normally be assigned its 
  # own interval. The code below takes the 31st day and joins it with the 
  # previous interval. 
  group_by(ID) %>% 
  mutate(new = if_else(day(new) == 31, new - days(10), new)) %>% 
  group_by(new, .add = TRUE) %>%
  # Filter the data by the season based on the `season_categ` column
  filter(month == "1") %>% 
  group_split()
```

### Intervals for July
```{r July Intervals}
july <- deer %>%
  # Creates a new column assigning the first day in the 10-day interval in which
  # the date falls under (e.g., 01-03-2021 would be in the first 10-day interval
  # so the `floor_date` assigned to it would be 01-01-2021)
  mutate(new = floor_date(date, "10 days")) %>%
  # For any months that has 31 days, the 31st day would normally be assigned its 
  # own interval. The code below takes the 31st day and joins it with the 
  # previous interval. 
  group_by(ID) %>% 
  mutate(new = if_else(day(new) == 31, new - days(10), new)) %>% 
  group_by(new, .add = TRUE) %>%
  # Filter the data by the season based on the `season_categ` column
  filter(month == "7") %>% 
  group_split()

```

Elements in the list for one individual in January:
```{r}
head(jan[1:3])
```


## Create Tracks for Each Interval
Before we estimate LoCoH home range for each of these 10-day intervals, we first create objects of class `track_xyt` using `make_track()` from the `amt` package. This function creates a `track` using the `utm_x` and `utm_y` columns from `jan` and `july`. `lapply()` allows us to apply the specified function across a `list` object (e.g., `jan` and `july`)

``` {r tracks, message = FALSE}
track_list_sum <- lapply(jan, function(x) {
  amt::make_track(tbl = x, .x = utm_x, .y = utm_y, .t = date,
             uid = ID,
             # lat/long: 4326 (lat/long, WGS84 datum).
             # utm: crs = sp::CRS("+init=epsg:32612"))
             crs = 32612)
})
track_list_win <- lapply(july, function(x) {
  amt::make_track(tbl = x, .x = utm_x, .y = utm_y, .t = date,
             uid = ID,
             crs = 32612)
})


```

## Creates LoCoHs Using the Tracks 
Using the `track` created for the `jan` and `july`, we estimate type "a" LoCoH home range using `hr_locoh()` from the `amt` package. If you would like to learn more about the arguments of this function please refer to `?amt::hr_locoh` for more details.  

```{r LoCoHs}
sum_locoh <- lapply(track_list_sum, function(tracks){
    # Calculate LoCoH a*
    dmat <- dist(tracks[, c("x_", "y_")])
    a <- max(dmat)

    # Fit LoCoH
    locoh <- amt::hr_locoh(
      x = tracks,
      levels = seq(0.1, 1, by = 0.1),
      keep.data = TRUE,
      n = a,
      type = "a",
      rand_buffer = 1e-05
    )
    
    
})
win_locoh <- lapply(track_list_win, function(tracks){
    # Calculate LoCoH a*
    dmat <- dist(tracks[, c("x_", "y_")])
    a <- max(dmat)

    # Fit LoCoH
    locoh <- amt::hr_locoh(
      x = tracks,
      levels = seq(0.1, 1, by = 0.1),
      keep.data = TRUE,
      n = a,
      type = "a",
      rand_buffer = 1e-05
    )
    
    
})
```

## Creates Isopleths from the LoCoHs
We then created isopleths from the LoCoH home range estimates using `hr_isopleths()`from the `amt` package. 
```{r Isopleths}
sum_iso <- lapply(sum_locoh, function(x){
  amt::hr_isopleths(x)
})
win_iso <- lapply(win_locoh, function(x){
  amt::hr_isopleths(x)
})

```

## Rasterizes the Isopleths 
After creating isopleths for each of the intervals, we have to rasterize them. Similiar to the the previous functions, we use `lapply()` to apply our function across our `list` object. We first create a template raster with a resolution of "50" (keep in mind that the speed the isopleths are `rasterize()` will be dependent on the specified resolution, with time of processing increasing with smaller resolutions). The projection of the template raster will match that of the specified isopleth.

After creating rasters for each interval, we then subtract all the isopleth values from, then normalize them. 
```{r, results = "hide"}
sum_raster <- lapply(sum_iso, function(iso){
# Creates template raster
    template_raster <- raster::raster(iso,
      resolution = 50, vals = 0,
      crs = sp::CRS(sf::st_crs(iso)[[2]])
    )
    # Rasterizes the isopleth
    raster_locoh <- raster::rasterize(iso,
      template_raster,
      field = "level",
      fun = "first"
    )

    # Now you want to subtract all the isopleth values from 1
    raster::values(raster_locoh) <- 1 - raster::values(raster_locoh)

    # Now normalize
    x <- raster_locoh / raster::cellStats(raster_locoh, sum)
})

win_raster <- lapply(win_iso, function(iso){
# Creates template raster
    template_raster <- raster::raster(iso,   
      resolution = 250, vals = 0,
      crs = sp::CRS(sf::st_crs(iso)[[2]])
    )
    # Rasterizes the isopleth
    raster_locoh <- raster::rasterize(iso,
      template_raster,
      field = "level",
      fun = "first"
    )

    # Now you want to subtract all the isopleth values from 1
    raster::values(raster_locoh) <- 1 - raster::values(raster_locoh)

    # Now normalize
    x <- raster_locoh / raster::cellStats(raster_locoh, sum)
})
```

## Earth Mover's Distance
The earth mover's distance (EMD), also known as the Wasserstein metric, is a measure of the distance between two probability distributions over a region. For example, if the distributions are two different ways of piling up a certain amount of dirt over a region, the EMD is the minimum effort needed in turning one pile into the other, where the cost is assumed to be the amount of dirt moved times the distance it is moved. 

## EMDs in Geographical Space

Calculates the EMD's between the summer and winter rasters in a geographical space. The `3` specifies that there are 3 elements(in our example 3 10-day intervals) per individual in a given season, this will change based on the length of the time frame selected for the intervals. 
```{r between geographical space EMDs}
btwgeoemd <- between_emd(sum_raster, win_raster, 3)

```

Calculates the EMD's between the intervals within a single season. The `3` specifies that there are 3 intervals for each individual within a single season. 
```{r Within geographical space EMDs}

with_sumgeo <- within_emd(sum_raster, 3)
with_wingeo <- within_emd(win_raster, 3)

```

## EM-Speeds
Calculate ‘EM-speeds’ by dividing each EMD by the time elapsed between the two UDs it corresponds to. For example, to calculate the EM-speeds between the first intervals in `jan` and `july`, the EMDs for the intervals are divided by the the difference in Julian dates between the starts dates of each interval. 

$EMD /(Julian.start.2 - Julian.start.1) $

We calculated the EM-speeds between seasons (e.g., first intervals between `jan` and `july`), and within seasons (e.g., first and second intervals between `jan` and `july`) using the `emspeeds()` in our `trekR` package.
```{r}
btwesp <- emspeeds(july, jan, btwgeoemd)

wesp_sum <- emspeeds(july, july, with_sumgeo)
wesp_win <- emspeeds(jan, jan, with_wingeo)
```

## Average Within and Between Season EM-Speeds
We calculated the average between-season and within-season EM-speeds. If an animel does not shift its space use between saeson, the within-season EM-speeds should be similar to the between-seasons EM-speeds. 
```{r}
btwavg_geo <- lapply(btwesp, mean)
wavg_geo_sum <- lapply(wesp_sum, mean)
wavg_geo_win <- lapply(wesp_win, mean)
```

The ratio between the two averages are the migratory predictors and are limitless. 

```{r}

btwavg_geo[[1]] / wavg_geo_sum[[1]]
btwavg_geo[[1]] / wavg_geo_win[[1]]

```
## EMDs in Environmental Space

### Creating 'RasterStacks' to calculate EMD and 'EM-Speeds in Environmental Space

We first import the environmental covariate file (e.g., DEM layer), then we reproject the extent from latitude and longitude to UTMs to match those of the home range rasters.
```{r}
  #Load raster
  UtahDEM <- raster::raster(system.file("extdata", "Utah_DEM.tif", package = "trekR"))
  out_proj <- "+proj=utm +zone=12 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"
  UtahDEM@extent <- raster::extent(raster::projectExtent(UtahDEM, out_proj))
```

The `env_raster_stack()` function takes the environmental covariate raster and resamples it to the home range rasters. It then takes the resampled rasters and the home range rasters and constructs a object with class `RasterStack`. The steps following the creation of these `RasterStacks` is the steps under "Earth Mover's Distance", where the `RasterLayers` objects (i.e., `sum_raster` and `win_raster`) are replaced with the `RasterStack` objects (i.e., `sum_rst` and `win_rst`). 

```{r}
sum_rst <- env_raster_stack(UtahDEM, sum_raster)
win_rst <- env_raster_stack(UtahDEM, win_raster)

```

